\chapter{Quotes About Statistics} \label{Quotes About Statistics}

Whilst not necessary for the study or understanding of statistics, below are some quotations about statistics (and mathematics in general) which the author has come across over the years.  They may amuse, inform or lead the student to introspection and consideration of the role of mathematics and statistics in the world.

\woptionrule{}

\begin{quote}
    I can prove anything with statistics except the truth \textemdash George Canning, Politician (1770-1827) \index[people]{Canning, George}
\end{quote}

\begin{aside}
This is often misquoted as \emph{``You can prove anything with statistics''}
\end{aside}

\cenoptionrule{}

\begin{quote}
    There are three kinds of lies: lies, damned lies, and statistics \textemdash Benjamin Disraeli, Politician (1804 - 1881) \index[people]{Disraeli, Benjamin}
\end{quote}

\begin{aside}
A similar quote (\emph{``Lies, damned lies and statistics''}) is often ascribed to Samuel Clemens.  \index[people]{Clemens, Samuel Langhorne}  It is unclear if either Disraeli or Clemens were quoting the other or had independently come up with the quote.
\end{aside}

\cenoptionrule{}

\begin{quote}
    Never tell me the odds \textemdash Han Solo, ``Star Wars: A New Hope''
\end{quote}

\cenoptionrule{}

\begin{quote}
    One death is a tragedy, a million deaths, a statistic \textemdash Josef Stalin, Soviet Leader, (1879 - 1953) \index[people]{Stalin, Joseph Vissarionovich} \emph{Attributed by The Washington Post, 20th January 1947}
\end{quote}

\begin{aside}That Stalin actually said this has been disputed a number of times.  Whilst it seems to be be very much inline with the zeitgeist of the USSR (United Soviet Socialist Republic) during the Second World War it is possibly worth bearing in mind that at the time the attribution was printed the USA was just entering the early stages of McCarthyism and trying to distance itself from the Communist allies of that war.\end{aside}
\cenoptionrule{}
\begin{quote}
    The good Christian should beware the mathematician and all those who make empty prophecies. The danger already exists that the mathematicians have made a covenant with the devil to darken the spirit and to confine man in the bonds of hell. \textemdash St Augustine of Hippo (354 - 430) \index[people]{Saint Augustine of Hippo}\index[people]{Hipponensis, Aurelius Augustinus }
\end{quote}

\cenoptionrule{}

\begin{quote}
    Facts are stubborn things, but statistics are pliable. \textemdash Samuel Clemens, writing as Mark Twain (1835 - 1910)\index[people]{Twain, Mark} \index[people]{Clemens, Samuel Langhorne}
\end{quote}

\cenoptionrule{}

\begin{quote} \label{qt:AndersonsLaw}
    I have yet to see any problem, however complicated, which, when you looked at it in the right way, did not become still more complicated. \textemdash ``Call me Joe'',  Poul Anderson (1926 - 2001)\index[people]{Anderson, Poul}\index[terms]{Anderson's Law}
\end{quote}



\optionrule{}









